---
title: "Practice tidyverse functions for data wrangling"
author: "Felicia Cruz"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Warm-up
```{r}
us_rent_income <- us_rent_income

us_rent_income_wider <- us_rent_income %>% 
  pivot_wider(names_from = variable,
              values_from = estimate) %>% 
  select(-moe)
```

## Practice tidyverse functions for data wrangling 

- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- piping thihngs together in sequence 

### dplyr::filter() to keep or exclude rows based on our conditions 

1. Exact match (single condition)

Only keep observations where the island is Biscoe 
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```

2. "and" statement within filter

Only keep observations if the penguin species is Chinstrap AND the year is 2008 
```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap" & year == 2008)

```

3. Exclude Gentoos 

Keep observations if the penguin species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")

```

Practice:
Create a subset from `penguins` that only contains observations for Adelie penguins with a flipper length greater than or equal to 185 mm.

```{r}
practice_1 <- penguins %>% 
  filter(species == "Adelie" & flipper_length_mm >= 185)

```

4. The `%in%` operator

  Use the %in% operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row.
  
  We want to keep observations where the island is Dream or Torgersen. 
  
```{r}
ex_4 <- penguins %>%  
  filter(island %in% c("Dream", "Torgersen")) 

```
Why couldn't you just use `==` instead of `%in%`? 
  It will look for the islands in that order and will keep repeating that process. 
  

### `select()` to select or exclude columns

5. Keep a single column by name 

```{r}
ex_5 <- penguins %>%  select(species)

# Rewrite without %>%  

ex_5 <- select(penguins, species)
```

6. Keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year
```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)

# could use `names()` to make sure the correct variables are returned
```

7. Choose or exclude columns based on position
```{r}
ex_7 <- penguins %>% 
  select(-4)

```

8. A piped sequence 

Write a piped sequence where you first: only keep columns for species, body_mass_g, and island, and second: filter to only include observations if body_mass_g is greater than 3600

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600) 
```

### `mutate()` to add or update existing columns, while keeping the ones already there

9. Add a column that has body mass converted to kg 
```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>%  
  mutate(flipper_length_m = flipper_length_mm / 1000)
```
You can also have multiple mutate statements inside one `mutate()` function OR can separate into multiple lines. 

10. Mutate to update existing column
```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HOORAY!"))
```
Can be used if somebody misspelled something the same way multiple times. Be careful because this overwrites an existing column. Can always just make a new column. 

11. Update the class of the bill_depth column to character
```{r}
ex_11 <- penguins %>%  
  mutate(bill_depth_mm = as.character(bill_depth_mm))

# to check class:
class(ex_11$bill_depth_mm)
```

### `case_when()`: your favorite if-else

12. Create a new column that contains "small" if penguin flipper length is less than 185 mm, "medium" otherwise
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))

```

Task: Starting from penguins, create a new column called "bird_bill_bin" that contains: 
  - If bill length < 38 mm: "short bill"
  - If bill length >= 38 mm & < 43 mm: "medium bill" 
  - If bill length >= 43: "long bill"
  
```{r}
practice_2 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
  ))

# You can also use `TRUE ~ "long bill"` as a "catch all" to apply something to everything else that does not fall into the previous categories. 
```

### `group_by() %>%  summarize()` to find summary values by group and return them in a nice summary table 

13. use `group_by()` to say there are groups within the species column
```{r}
ex_13 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), 
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

# can add as many different groupings as we want 
```

## Practice 

In a single piped sequence: 
- Only keep penguins with a body mass > 3500 g
- Only keep columns for species, flipper_length_mm, and sex
- Create a summary table that contains the mean, sd, max, and min value of flipper length by species and sex (as the grouping variables)

```{r}
practice_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flipper_length = max(flipper_length_mm, na.rm = TRUE),
            min_flipper_length = min(flipper_length_mm, na.rm = TRUE)) %>% 
  ungroup()


```


